International Relations are a Social Network:\\ A Latent Network Approach for Global Politics

### 2017-10-20 paper/4_app.tex
interim dock
yeah, I know I should fork, but why fix my manners now?
out and synced

We choose five prominent studies from the broad field of international relations and international political economy that utilize relational data \citep{reiter:stam:2003, mcdonald:2004,  rose:2004, weeks:2012, gibler:2017}. These studies are recent and have been cited over 100 times. Each of these pieces was published in a prominent journal and is well-known in the literature. Each used the standard approach in political science, which is to employ some form of a general linearized regression that ignores dyadic interdependencies, except as they may reveal themselves in included variables. Also part of the gold standard is a post estimation cleanup thought to be a paliative, and each of these studies adjust the posterior standard errors in an attempt to account for the clustering of observations.

@@

Given current practices, most scholars have a single variable in a complicated empirical model that they look at for evaluating the validity of their empirically estimated models.  Whether putative control variables, for example, are precisely estimated is less important. Generally, interpretations focus on a small set of independent variables.  In the study by Reiter and Stam (2003) the main question is to determine in mixed dyads---consisting of a democratic country and country ruled by a personalist dictatorship---whether it is the
democratic country or the personalist country that is associated with a greater likelihood that the dyad is in a dispute.  In a regression including fourteen coefficients, they focus on whether the variables that capture whether the initiator is a democratic country are statistically significant.\footnote{In particular, the variables \texttt{Pers/Democ} and \texttt{Democ/Pers} in Model 3, in Table 1, page 335.} In our evaluation we focus on the same question in the re-examination of the Reiter and Stam study to determine whether the model replicates when reestimated within the AME framework. Each of the other four studies also has a crucial finding that we hone in on to further draw into focus the potential differences brought about by utilizing an AME estimation procedure.  In Table~\ref{tab:modelFindingSumm}, we present the overall results; the term \textbf{fails to replicate} indicates only that the sign and/or significance of the crucial finding in the original study is not also found in the AME estimation.

@@

\begin{table}[ht]
\centering
	\begin{tabular}{l|l cc}
	~ & ~ & GLM & AME \\
	\toprule
\multirow{2}{*}{Reiter \& Stam (2003)} & AUC (ROC) & 0.92 &   {0.96} \\
		~ & AUC (PR) & 0.08 &  {0.15} \\		\midrule
\multirow{2}{*}{McDonald (2004)} & AUC (ROC) & 0.92 &  {0.99} \\
		~ & AUC (PR) & 0.13 &  {0.28} \\		\midrule
\multirow{2}{*}{Rose (2004)} & $ (1/n \sum \varepsilon^2)^{1/2} $ (RMSE) & 3.23 &  {1.99} \\
		~ & $(\text{median}\quad (\varepsilon^2))^{1/2}$ (RMDSE) & 2.01 &  {1.06} \\	\midrule
\multirow{2}{*}{Weeks (2012)} & Area Under ROC Curve (AUC-ROC) & 0.64 &  {0.97} \\
		~ & Area Under PR Curve (AUC-PR) & 0.00 &  {0.15} \\		\midrule
\multirow{2}{*}{Gibler (2017)} & AUC (ROC) & 0.52 &  {0.91} \\
		~ & AUC (PR) & 0.00 &  {0.08} \\			\bottomrule
	\end{tabular}
	\caption{Here we provide a summary of the out-of-sample performance based on our cross-validation strategy for each of the five replications when using the standard dyadic approach and our network based approach. Four of the five studies involved a binary dependent variable, so for those measures we provide area under the curve (AUC) statistics. The Rose study involved a Gaussian dependent variable and for that we use the root mean squared error (RMSE) and root median squared error (RMDSE). Cases in which our network based approach outperformed the standard approach are highlighted in bold.}
	\label{tab:modelPerfSumm}
\end{table}
\FloatBarrier

@@

\citet{mcdonald:2004} studies whether trade promotes peace between nations. The link between conflict and trade is perplexing, with many presistent, yet competing explanations. \citet{mcdonald:2004} includes the argument that interdependence between states ``makes conflict less likely because of its efficiency over conquest in acquiring resources\ldots (547)''. Accordingly, his primary contribution is to provide evidence challenging the generalized linkage between peace and trade and to offer a new measurement of the key independent variable, trade. \citet{mcdonald:2004} refines the trade variable, arguing that \textit{free} trade, rather than trade alone, reduces the likelihood of conflict between states. His key hypothesis that greater levels of protection increase the probability of interstate conflict, an argument that builds on the work of classic liberalism and connects free trade to the power of domestic audiences. \citet{mcdonald:2004} measures free trade in two ways. The first captures the idea that larger protected sectors generate greater societal pressures resulting in pockets of support for war. This protection variable measures the proportion of customs revenue divided by total imports in the state that possesses the greater such ratio in each dyad. This measure captures the score of the state in the dyad that possesses higher barriers to trade (560). \citet{mcdonald:2004} also includes a measure of economic integration  calculated as ``the lower proportion of total dyadic trade (imports plus exports) divided by state $i$'s GDP or total dyadic trade divided by state $j$'s GDP'' (560). The binary, dependent variable is the onset of a new militarized interstate dispute within a given dyad. \citet{mcdonald:2004} employs logistic regression to examine the putative statistical significance of these variables. The models include splines to correct for temporal dependence, and robust standard errors clustered on each dyad.

@@

%\begin{figure}
%\includegraphics[width=\textwidth]{rose_coefs_all_final.pdf}
% \caption{Coefficient plot of Rose (2004)'s original model (blue) compared to AME model (red).}\label{fig:roseCoef}
%\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{rose_aeff_top10.pdf}
 \caption{Nodal Random Effects for AME estimation of Rose (2004)}\label{fig:roser}
\end{figure}
\FloatBarrier
%\begin{figure}
%\centering   
%\subfigure[Nodal Random Effects]%{\label{fig:rosesend}\includegraphics[width=65mm]{ABplot_rose_top10.pdf}}
%\subfigure[Receiver Random Effects]{\label{fig:rosereceive}\includegraphics[width=65mm]{ABplot_receiver_rose.pdf}}
%\caption{Sender and receiver random effects for AME estimation of Rose (2004).}
%\label{fig:rosesr}
%\end{figure}

@@

%. THINK THIS IS GONE 
%\begin{figure}
%	\includegraphics[width=\textwidth]{weeks_coefs_all.pdf}
%	\caption{\label{fig:weeksCoef} Coefficient plot of Weeks' (2012) original model (blue) compared to AME model (red). }
%\end{figure}
\begin{figure}
	\includegraphics[width=\textwidth]{weeks_circPlot.pdf}
	\caption{\label{fig:weekscirc} Visualization of multiplicative effects for Weeks (2012). Blue represents groups with common sending patterns and red represents groups with common receiving patterns.}
\end{figure}

@@

% We chose this article because it is published in 2017, demonstrating that the concern with interdependence continues to the present and because it was published in the so-called flagship journal of political science, \textit{The American Political Science Review.} It examines the link between dyadic capability differences and international conflict in terms of how and when states became independent members of the international system.  It assumes that membership in the international system---defined in typical ways---has a substantial influence on capabilities. In the same way that capabilities don't seem to vary enormously over time, at least comparatively, states become independent in a particular location, clustered with surrounding states. Neither does this change much over time. Gibler shows that the long-standing relationship between the relative parity of capabilities and initiation of international conflict is almost completely mediated by the initial conditions for the members of the dyad when they joined the international system as sovereign members. This finding calls into question many IR theories about the role of balance in terms of generating international conflict \citep{organski:1958}.

@@

\begin{table}
\begin{center}
\caption{Comparison of Gibler (2017) Model 6 results with AME results. This was run over a chain of $10,000$ iterations.  \label{tab:gibme}}
\begin{tabular}{lrr} \toprule
& \multicolumn{2}{c}{$\hat{\beta}$ Estimates}\\ \cmidrule{2-3}
Variable & \texttt{logit} & \texttt{amen} \\ \midrule
Allied & 0.142&  0.023 \\
Joint Democracy &\bf -0.507&  0.045 \\
Peace Years \bf &-0.260& \bf -0.060 \\
Spline 1 &\bf -0.001& \bf 0.00 \\
Spline 2&\bf -0.000& \bf 0.00\\
Spline 3 &-0.000& \bf 0.00\\ 
Contiguity &\bf 2.412& \bf 0.640 \\
Parity &0.075&  -0.013 \\
Parity at entry year&\bf 0.868&  0.002 \\ 
Rivalry &\bf 2.031&\bf 0.721\\ 
Constant &\bf -5.526 & \bf  -2.581\\ \bottomrule
\end{tabular}\\
\end{center}
{\bf Note}: {\bf Bold} indicates conventional statistical significance at $p < 0.05$ or less in Gibler (2017). For comparative purposes, only, we have employed the same criterion to the results from the \texttt{amen} estimation.
\end{table}

We re-estimated model 6 from Table 6 (2017,34). The results are presented in Table~\ref{tab:gibme} in order to facilitate explicit comparison.
The results obtained with \texttt{amen} stand in stark contrast to those found with a logistic regression (with dyad clustered, robust standard errors).  Most importantly, the primary variable from the Gibler study, parity of the members of a dyad at the year in which they entered the international system, is shown to be unimportant in the amen results.  Not only is the value of the this parameter small, but it has a very large relative standard error, over a magnitude larger than the parameter itself ($z= 0.038$). In addition, joint democracy follows the same pattern of importance in the logistic results, but this disappears once interdependencies are modeled.  As might be expected the strong geographic clustering in the original study is about one-quarter as strong in the \texttt{amen} estimations. Similarly, rivalry coefficients are about one-third the size in the \texttt{amen} formulation, but a great deal more precisely measured ($z=18.116$).

Beyond more informative fixed effect coefficients, the \texttt{amen} approach also provides information about the interdependencies that were modeled. The most pertinent of these may be the dyadic and triadic dependencies which are presented in Figure~\ref{fig:gibler:dyadtriad}, which further reveal that the assumption of independence among the dyadic data in this study can be strongly rejected. Perhaps most importantly, the main substantive conclusions of the 2017 Gibler study do not seem warranted from the perspective of the results obtained with the \texttt{amen} estimation that explicitly models \first- \second-, and \third-order interdependencies. Not only do the estimated coefficients tell a different story, but the assumptions under which the original results would hold are shown to be violated by the data.

% MDW to SM:  yeah, Rose has so many variables that the network effects aren't that strong.
% Maybe we should mention this. If you put in a gazillion variables you might be okay.
% SM to MW: hmmm we should talk about the utility of this.
% see fig gibler_srmvc.pdf in the paper dir
% \begin{figure}
% 	\caption{}
% 	\includegraphics[width=\textwidth]{gibler_srmvc.pdf}
% 	\label{fig:gibler:dyadtriad}
% \end{figure}
% \FloatBarrier

@@

Third, it is clear that the AME approach dominates the OLS and Logit-based approaches in terms of performance. Not only it is better at correctly identifying cases in which the dependent variable takes a value of $0$ (via the ROC curves and associated statistics), but it also dominates at correctly identifying occurrences of the dependent variable in the data (seen via the PR curves and associated statistics).  In the case of studies
with continuous dependent variables, the AME approach has average error statistics that are about one-half that found in the OLS model. It is rare for
studies in this area to provide performance statistics, but at the same time that at least one of the studies is unable to identify a single case in spite of having almost a million observations.

Fourth, it is useful to suggest that a lot of what is \textit{known} about international relations based on studies of dyadic data should be taken
with a grain of salt, and awaits a reassessment using methods that do not rely on an assumption that all the dyads are
independent and identically distributed.

>>>>

We choose five prominent studies from the broad field of international relations and international political economy that utilize relational data \citep{reiter:stam:2003, mcdonald:2004,  rose:2004, weeks:2012, gibler:2017}. These studies are recent and have been cited over 100 times. Each of these pieces was published in a prominent journal and is well-known in the literature. Each used the standard approach in political science, which is to employ some form of a general linearized regression that ignores dyadic interdependencies, except as they may reveal themselves in included variables. Also part of the gold standard is a post estimation cleanup thought to be a palliative, and each of these studies adjust the posterior standard errors in an attempt to account for the clustering of observations.

@@

Given current practices, most scholars have a single variable in a complicated empirical model that they look at for evaluating the validity of their empirically estimated models.  Whether putative control variables, for example, are precisely estimated is less important. Generally, interpretations focus on a small set of independent variables.  In the study by Reiter and Stam (2003) the main question is to determine in mixed dyads---consisting of a democratic country and country ruled by a personalist dictatorship---whether it is the
democratic country or the personalist country that is associated with a greater likelihood that the dyad is in a dispute.  In a regression including fourteen coefficients, they focus on whether the variables that capture whether the initiator is a democratic country are statistically significant.\footnote{In particular, the variables \texttt{Pers/Democ} and \texttt{Democ/Pers} in Model 3, in Table 1, page 335.} In our evaluation we focus on the same question in the re-examination of the Reiter and Stam study to determine whether the model replicates when re-estimated within the AME framework. Each of the other four studies also has a crucial finding that we hone in on to further draw into focus the potential differences brought about by utilizing an AME estimation procedure.  In Table~\ref{tab:modelFindingSumm}, we present the overall results; the term \textit{does not replicate} indicates only that the sign and/or significance of the putatively crucial finding in the original study is not also found in the AME estimation.

@@

\begin{table}[ht]
\centering
	\begin{tabular}{l|l cc}
	~ & ~ & GLM & AME \\
	\toprule
\multirow{2}{*}{Reiter \& Stam (2003)} & Area Under ROC Curve, AUC-ROC & 0.92 & {0.96} \\
		~ & Area Under PR Curve, AUC-PR & 0.08 &  {0.15} \\		\midrule
\multirow{2}{*}{McDonald (2004)} & AUC-ROC & 0.92 &  {0.99} \\
		~ & AUC-PR & 0.13 &  {0.28} \\		\midrule
\multirow{2}{*}{Rose (2004)} & Root Mean Squared Error RMSE & 3.23 &  {1.99} \\
		~ & Root Median Squared Error RMDSE & 2.01 &  {1.06} \\	\midrule
\multirow{2}{*}{Weeks (2012)} & AUC-ROC & 0.64 &  {0.97} \\
		~ & AUC-PR & 0.00 &  {0.15} \\		\midrule
\multirow{2}{*}{Gibler (2017)} & AUC-ROC & 0.52 &  {0.91} \\
		~ & AUC-PR & 0.00 &  {0.08} \\			\bottomrule
	\end{tabular}
	\caption{Here we provide a summary of the out-of-sample performance based on our cross-validation strategy for each of the five replications when using the standard dyadic approach and our network based approach. Four of the five studies involved a binary dependent variable, so for those measures we provide area under the curve (AUC) statistics. The Rose study involved a Gaussian dependent variable and for that we use the root mean squared error (RMSE) and root median squared error (RMDSE). Cases in which our network based approach outperformed the standard approach are highlighted in bold.}
	\label{tab:modelPerfSumm}
\end{table}
\FloatBarrier

@@

\citet{mcdonald:2004} studies whether trade promotes peace between nations. The link between conflict and trade is perplexing, with many persistent, yet competing explanations. \citet{mcdonald:2004} includes the argument that interdependence between states ``makes conflict less likely because of its efficiency over conquest in acquiring resources\ldots (547)''. Accordingly, his primary contribution is to provide evidence challenging the generalized linkage between peace and trade and to offer a new measurement of the key independent variable, trade. \citet{mcdonald:2004} refines the trade variable, arguing that \textit{free} trade, rather than trade alone, reduces the likelihood of conflict between states. His key hypothesis that greater levels of protection increase the probability of interstate conflict, an argument that builds on the work of classic liberalism and connects free trade to the power of domestic audiences. \citet{mcdonald:2004} measures free trade in two ways. The first captures the idea that larger protected sectors generate greater societal pressures resulting in pockets of support for war. This protection variable measures the proportion of customs revenue divided by total imports in the state that possesses the greater such ratio in each dyad. This measure captures the score of the state in the dyad that possesses higher barriers to trade (560). \citet{mcdonald:2004} also includes a measure of economic integration  calculated as ``the lower proportion of total dyadic trade (imports plus exports) divided by state $i$'s GDP or total dyadic trade divided by state $j$'s GDP'' (560). The binary, dependent variable is the onset of a new militarized interstate dispute within a given dyad. \citet{mcdonald:2004} employs logistic regression to examine the putative statistical significance of these variables. The models include splines to correct for temporal dependence, and robust standard errors clustered on each dyad.

@@

\begin{figure}
\includegraphics[width=\textwidth]{rose_aeff_top10.pdf}
 \caption{Nodal Random Effects for AME estimation of Rose (2004)}\label{fig:roser}
\end{figure}
\FloatBarrier

@@

\begin{figure}
	\includegraphics[width=\textwidth]{weeks_circPlot.pdf}
	\caption{\label{fig:weekscirc} Visualization of multiplicative effects for Weeks (2012). Blue represents groups with common sending patterns and red represents groups with common receiving patterns.}
\end{figure}

@@

@@

THIS TABLE WILL BECOME A FIGURE, and skedaddle.

\begin{table}
\begin{center}
\caption{Comparison of Gibler (2017) Model 6 results with AME results. This was run over a chain of $10,000 \text{BAZILLION}$ iterations.  \label{tab:gibme}}
\begin{tabular}{lrr} \toprule
& \multicolumn{2}{c}{$\hat{\beta}$ Estimates}\\ \cmidrule{2-3}
Variable & \texttt{logit} & \texttt{amen} \\ \midrule
Allied & 0.142&  0.023 \\
Joint Democracy &\bf -0.507&  0.045 \\
Peace Years \bf &-0.260& \bf -0.060 \\
Spline 1 &\bf -0.001& \bf 0.00 \\
Spline 2&\bf -0.000& \bf 0.00\\
Spline 3 &-0.000& \bf 0.00\\ 
Contiguity &\bf 2.412& \bf 0.640 \\
Parity &0.075&  -0.013 \\
Parity at entry year&\bf 0.868&  0.002 \\ 
Rivalry &\bf 2.031&\bf 0.721\\ 
Constant &\bf -5.526 & \bf  -2.581\\ \bottomrule
\end{tabular}\\
\end{center}
{\bf Note}: {\bf Bold} indicates conventional statistical significance at $p < 0.05$ or less in Gibler (2017). For comparative purposes, only, we have employed the same criterion to the results from the \texttt{amen} estimation. NEED TO BE FIXED if staying.
\end{table}

We re-estimated model 6 from Table 6 (2017, 34). The results are presented in Table~\ref{tab:gibme} in order to facilitate explicit comparison.
The results obtained with \texttt{amen} stand in stark contrast to those found with a logistic regression (with dyad clustered, robust standard errors).  Most importantly, the primary variable from the Gibler study, parity of the members of a dyad at the year in which they entered the international system, is shown to be unimportant in the AME results.  Not only is the value of this parameter small, but it has a very large relative standard error---over a magnitude larger than the parameter itself ($z= 0.038$). In addition, the variable indicating whether both members of the dyad were coded as democracies (joint democracy) follows the same pattern: important and strong in the logistic results, but this disappears once interdependencies are modeled.  As might be expected, the strong geographic clustering in the original study is about one-quarter as strong in the AME estimations. Similarly, rivalry coefficients are about one-third the size in the AME formulation, but a great deal more precisely measured ($z=18.116$).

We utilized the original and the AME results from Gibler's model 6 in Table 6 (as above) to calculate the expected values for one scenario. We focused on the variables measuring rivalry.  We used mean or modal values for all independent variables, except we changed the rivalry variable to indicate that there was a rivalry when the actual data suggest there is none.  The expected values of this scenario are essentially a first difference plot comparing results with the model when estimated in two different ways: Gibler's GLM estimation and our AME approach.  As this Figure~\ref{fig:gibmargeff} illustrates, the AME results differ from the GLM in two distinct ways. First, the expected value of the dependent variable---the probability of the onset of a militarized interstate dispute, is considerably higher when taking interdependencies into account with the AME model.  These are rare events, so the probabilities are low, but the difference is a factor of $3x$. Thus, you get quite substantially different expected values from these two models.  Second, the GLM model has a standard errors which underestimate the true variability, in part because some of them are assumed to be zero.  This shows up in the narrow density of predicted outcomes for the GLM model, compared to the density seen with the AME. The latter reflects a wider range of potential outcomes, reflecting the incorporation of uncertainty in the AME model.

\begin{figure}
	\caption{}
	\includegraphics[width=\textwidth]{gibler_srmvc.pdf}
 	\label{fig:gibmargeff}
 \end{figure}

Beyond more informative fixed-effects coefficients, the AME approach also provides information about the interdependencies that were modeled. The assumption of independence among the dyadic data in this study can be strongly rejected. Perhaps most importantly, the main substantive conclusions of the 2017 Gibler study do not seem warranted from the perspective of the results obtained with the \texttt{amen} estimation that explicitly models \first- \second-, and \third-order interdependencies. Not only do the estimated coefficients tell a different story, but the assumptions under which the original results would hold are shown to be violated by the data.

@@

Third, it is evident that the actual results---not the estimated coefficents and their covariances---which are generated by the models differ greatly in expectations.  This implies that policy experimentations with the models, as well as scenario-based simulations and forecasting of GLM models are likely to often give misleading results compared to the AME approach.

Fourth, it is clear that the AME approach dominates the OLS and Logit-based approaches in terms of performance. Not only it is better at correctly identifying cases in which the dependent variable takes a value of $0$ (via the ROC curves and associated statistics), but it also dominates at correctly identifying occurrences of the dependent variable in the data (seen via the PR curves and associated statistics).  In the case of studies
with continuous dependent variables, the AME approach has average error statistics that are about one-half that found in the OLS model. It is rare for studies in this area to provide performance statistics, but at the same time that at least one of the studies is unable to identify a single case in spite of having almost a million observations.

Fifth, it is useful to suggest that a lot of what is \textit{known} about international relations based on studies of dyadic data should be taken with a grain of salt, and awaits a reassessment using methods that do not rely on an assumption that all the dyads are independent and identically distributed.
